Summer 2025 ToDo

### 2025-06-06 Relations/Paper/Definitions.tex
Paper revision

\subsection{Closure operators (\texttt{Closure.agda})}

\begin{notation}
The following notations are standard.
  \begin{enumerate}
    \item $R^r$ denotes the \emph{reflexive closure} of $R$.
    \item $R^s$ denotes the \emph{symmetric closure} of $R$.
    \item $\tclos{R}$ denotes the \emph{transitive closure} of $R$.
    \item $R^* = (R^r)\tclos{}$ denotes the \emph{reflexive-transitive closure} of $R$.
    \item $R^= = (R^s)^*$ denotes the \emph{equivalence relation generated by} $R$.
  \end{enumerate}
\end{notation}

@@

@@

  The following claims are formalized in \texttt{Implications.agda}.
    \begin{enumerate}
        \item \verb|NP₌↔NP : R isNP₌ ↔ R isNP|
        \item \verb|UN→|\texttt{UN}$^{\to}$\verb|: R isUN → R is|\texttt{UN}$^{\to}$
    \end{enumerate}
\end{proof}

@@

\subsection{Sequences and Recurrence (\texttt{Properties.agda}, \texttt{Seq.agda})}
% The following definitions are key to formalizing Theorem 1.2.3 in \terese.
The definitions in this subsections are key to formalizing Theorem 1.2.3 in \terese.

@@

Having a size function that maps all of $A$ to the natural numbers can be computationally taxing to satisfy.
Furthermore, it takes us outside of a purely relational theory. The following properties will allow us to substitute $\gInc$ with weaker hypotheses in all the relevant places.
(For further discussion, see Subsection \ref{subsec:def}.)

\begin{definition} \label{def:rp} Let $a \in A$ and $s : \nat \to A$.
    \begin{description}
        % \item[\ule{$s : \nat \to A$}] \emph{A sequence $s \in A$} is a function from the natural numbers to $A$.
        % \item[\ule{$s$ is $R$-increasing}] \emph{A sequence $s \in A$ is $R$-increasing} if every term is $R$-related to its preceding term.
        \item[\ule{$a$ is an $s$-bound}] if $s (i) \mstep a$ for all $i \in \nat$.
        \item[\ule{$a \in \BP_R$}] \emph{$a$ has the bound property} if for each $R$-increasing sequence $s$ with $s (0) = a$ there exists an $s$-bound.
        \item[\ule{$\gRP(R)$}] \emph{$R$ has the recurrence property} if for every bounded, increasing sequence $s$ with an $s$-bound, there exists an $i \in \nat$ such that
        $s (i) \mstep x \implies x \mstep s (i)$ for all $x\in A$.
        \item[\ule{$\gRPm(R)$}] \emph{$R$ has the weak recurrence property} if for every increasing sequence $s$ with bound $b$ there exists an $i \in \nat$ such that $b \mstep s (i)$.
    \end{description}
\end{definition}

In \terese $\mathrm{\,Ind}$ is used to denote $\BP$.

@@

\begin{proof}
    \begin{enumerate}
        \item Immediate.
        \item See \verb|RP-↔RP : R isRP- ↔ (R isRP)| in \texttt{Implications.agda}.
    \end{enumerate}
\end{proof}

%TODO: Update below to reflect new convenion for proofs and code.
\begin{remark}
    It follows that every relation that is increasing in the sense of \terese $\,$ vacuously satisfies the RP condition.
\end{remark}

>>>>

\subsection{Closure operators (\texttt{ClosureOperators.agda})}

\begin{notation}
The following notations are standard.
  \begin{enumerate}
    \item $R^r$ denotes the \emph{reflexive closure} of $R$.
    \item $R^s$ denotes the \emph{symmetric closure} of $R$.
    \item $\tclos{R}$ denotes the \emph{transitive closure} of $R$.
    \item $R^* = (R^r)\tclos{}$ denotes the \emph{reflexive--transitive closure} of $R$.
    \item $R^= = (R^s)^*$ denotes the \emph{equivalence relation generated by} $R$.
  \end{enumerate}
\end{notation}

@@

The file \texttt{ClosureOperators.agda} verifies a number of properties of these operators
on relations needed in the subsequent development, including interaction between them,
monotonicity, preservation of logical equivalence, etc.
\apcomment{Edit/Improve this!}

@@

  The following claims are formalized in \texttt{Implications.agda}.
    \begin{enumerate}
        \item \verb|NP₌↔NP : R isNP₌ ↔ R isNP|
        \item \verb|UN→|\texttt{UN}$^{\to}$\verb|: R isUN → R is|\texttt{UN}$^{\to}$ \qedhere
    \end{enumerate}
\end{proof}

@@

\subsection{Sequences and Recurrence (\texttt{Seq.agda})}
% The following definitions are key to formalizing Theorem 1.2.3 in \terese.
The definitions in this subsections are key to formalizing Theorem 1.2.3 in \terese.

\begin{definition}\hfill
    \begin{description}
        \item[\ule{$x \in CP_R$}] \emph{$x$ has the cofinality property} if there exists an $R^r$-increasing sequence
        \mbox{$s : \nat {\to} A$} starting from $x$ such that every
        reduct of $x$ reduces to some element of the sequence.
        \[ x \mstep y \implies \exists i. y \mstep s(i) \]
        % \item[\ule{$x \in CP_R$}] \emph{$x$ has the cofinality property} if there exists an
        % (not necessarily strictly) increasing sequence starting from $x$ such that every
        % reduct of $x$ reduces to some element of the sequence.
    \end{description}
\end{definition}

\begin{definition} \label{def:bound} \hfill
  \begin{description}
    \item[\ule{$a \in A$ is an $s$-bound}] if for all $i \in \nat$, $s (i) \mstep a$.
    \item[\ule{$s : \nat \to A$ is bounded}] if there exists $a \in A$ that is an $s$-bound.
  \end{description}
\end{definition}

\begin{definition} \label{def:rp}
  The following are global properties of the given relation $R$.  \hfill
    \begin{description}
        \item[\ule{$\gBP(R)$}] \emph{$R$ has the boundedness property} if every $R$-increasing sequence is bounded.
        \item[\ule{$\gRP(R)$}] \emph{$R$ has the recurrence property} if for every bounded, increasing sequence $s$ with an $s$-bound, there exists an $i \in \nat$ such that
        $s (i) \mstep x \implies x \mstep s (i)$ for all $x\in A$.
        \item[\ule{$\gRPm(R)$}] \emph{$R$ has the weak recurrence property} if for every increasing sequence $s$ with bound $b$ there exists an $i \in \nat$ such that $b \mstep s (i)$.
    \end{description}
\end{definition}

In \terese, $\mathrm{\,Ind}$ is used to denote $\gBP$.

\terese also defines the following property, which is not included in our formalization.

@@

Having a size function that maps all of $A$ to the natural numbers can be computationally taxing to satisfy.
Furthermore, it takes us outside of a purely relational theory. The weaker property $\gRP$ defined above will substitute $\gInc$ all places that require it. (For further discussion, see Subsection \ref{subsec:def}.)

@@

\begin{proof} \hfill
    \begin{enumerate}
        \item $\gInc$ implies that there are no bounded $R$-increasing sequences,
        so $\gRP$ holds vacuously.
        \item See \verb|RP-↔RP : R isRP- ↔ R isRP| in \texttt{Seq.agda}. \qedhere
    \end{enumerate}
\end{proof}

\apcomment{Make sure to move this stuff to Seq.agda!}

% %TODO: Update below to reflect new convenion for proofs and code.
% \begin{remark}
%     It follows that every relation that is increasing in the sense of \terese $\,$ vacuously satisfies the RP condition.
% \end{remark}

### 2025-06-06 Relations/Paper/Formalization.tex
Paper revision

\subsection{Formalization of Definitions}\label{subsec:def}
The closure operators are formalized in \texttt{Closure.agda}. 
The definitions grouped as confluent in \terese $\,$ are formalized in \texttt{Base.agda}. 
The definitions grouped as normalizing in \terese $\,$ are formalized in \texttt{Properties.agda}.

@@

\subsection{Propositions}
There are three sets of propositions given in \terese: \texttt{1.1.9}, \texttt{1.1.10}, \texttt{1.1.11}. 
The formalization of the proofs are in \texttt{Propositions.agda}. These formalizations are straightforward 
and as such are not the focus of this paper.

\subsection{Theorems}\label{subsec:theorems}
\subsubsection{Newman's Lemma}
There are three proofs of Newman's Lemma ($\SN \land \WCR \implies \CR$) given in \terese. 
The first and third of these proofs make use of classical reasoning. The second proof is amenable to a 
constructive formalization as shown in the following function in \texttt{ARS-NewmansLemma.agda}:

@@

We show in Subsection \ref{subsec:newnewman} a generalization of Newman's Lemma using a condition distilled from $\RP$. 
The formalizations of the remaining theorems are all found in \texttt{ARS-Theorems.agda}. 
\subsubsection{Theorem 1.2.2}
There are three parts to this theorem. 
\begin{description}
    \item[($i$)] $\gCR \implies \gNPe \implies \gUN$
    \item[($ii$)] $\gWN \land \gUN \implies \gCR$  
    \item[($iii$)] $\gCRs \implies \gCR$
\end{description}

\begin{proof}
    See, \\
    \verb|CR→NP : R isCR → R isNP₌|\\
    \verb|NP→UN : R isNP₌ → R isUN|\\
    \verb|ii : R isWN × R isUN → R isCR|\\
    \verb||\texttt{UN}$^{\to}$ \verb|∧WN→CR : R |\texttt{isUN}$^{\to}$ \verb| → R isWN → R isCR|\\
    \verb|iii : subcommutative R → R isCR|\\
    in \texttt{Theorems.agda}.
\end{proof}
\begin{remark}
    In $\mathtt{UN^{→}∧WN→CR}$ a more general proof of ($ii$) is given using $\UNto$.
\end{remark}

\subsubsection{Theorem 1.2.3}
The following definition is required for this theorem.
\begin{definition}\hfill
    \begin{description}
        \item[\ule{$a \in CP_R$}] $a$ has the \emph{cofinality property} if there exists an 
        (not necessarily strictly) increasing sequence starting from $a$ such that every
        reduct of $a$ reduces to some element of the sequence.
    \end{description}
\end{definition}

There are four parts to this theorem. 
\begin{description}
    \item[($i$)] $\gWN \land \gUN \implies \gBP$
    \item[($ii$)] $\gBP \land \gInc \implies \gSN$  
    \item[($iii$)] $\gWCR \land \gWN \land \gInc \implies \gSN$
    \item[($iv$)] $\gCR \iff \gCP$ ($\gCR \implies \gCP$ only holds if the ARS is countable.)
\end{description}
\begin{proof}
    See, \\
    \verb|i : R isWN → R isUN → R isBP|\\
    \verb|i→ : R isWN → R |\texttt{isUN}$^{\to}$ \verb|→ R isBP|\\
    \verb|iiSeq : R isWN → R isUN → R isRP → isWFseq- (~R R)|\\
    \verb|iii : R isWN → R isWCR → R isRP- → dec (SN) → R isSN|\\
    \verb|iv : R isCP → R isCR| \\
    in \texttt{Theorems.agda}.
\end{proof}

\begin{remark} \hfill
    \begin{itemize}        
        \item In $\mathtt{i\to}$ we again provide a slight improvement by using $\gUNto$ rather than $\gUN$.
        \item ($ii$) is inherently classical. In $\mathtt{iiSeq}$, we show that these properties imply $\isWFseqm$; establishing the implication to $\gSN$ would require additional classical assumptions. The definition and further discussion of $\isWFseqm$ are provided in Section~\ref{sec:Well-foundedness}. 
        \item In $\mathtt{iii}$, we replace $\gInc$ with the more general property $\gRPm$. We also assume that $\gSN$ is decidable, as this allows the construction of a witness for whether an element possesses the property $\SN$, which is essential for our proof.
        \item We did not formalize $\gCR \to \gCP$ as it requires cardinality assumptions.
    \end{itemize}
\end{remark}

>>>>

\subsection{Formalization of Definitions (\texttt{Base.agda})}\label{subsec:def}

\apcomment{Rephrase this?  Or rewrite this subsection.}

Most standard termination and confluence properties are defined in the file \texttt{Properties.agda}
and related files discussed in the previous seciton.  The file \texttt{Base.agda} imports all of these files
and introduces a few more notations used specifically for formalizing chapter 1 of \cite{Terese}.
\apcomment{Discuss these...}

The closure operators are formalized in \texttt{Closure.agda}.
The definitions grouped as confluent in \terese $\,$ are formalized in \texttt{Base.agda}.
The definitions grouped as normalizing in \terese $\,$ are formalized in \texttt{Properties.agda}.

\apcomment{Should articulate the fact that normalization is defined as well-foundedness of the \emph{converse}
relation, per \cite{Terese}.  Will discuss this separately in WF section, but should remind the reader of this
whenever SN/SM comes up. }

\apcomment{When SMseq comes up, also point to section 5, and the need for classical logic in going from
sequential versions to inductive ones.  IOW, elaborate the summary below. }

@@

\subsection{\texttt{Propositions.agda}}
There are three sets of propositions given in \terese: \texttt{1.1.9}, \texttt{1.1.10}, \texttt{1.1.11}.
The formalization of the proofs are in \texttt{Propositions.agda}. These formalizations are straightforward
and as such are not the focus of this paper.

\subsection{\texttt{Theorems.agda}} \label{subsec:theorems}

\begin{theorem}(Newman's Lemma)
  $\gSN \land \gWCR \implies \gCR$
\end{theorem}

\begin{proof}
  There are three proofs of Newman's Lemma given in \terese.
  The first and third of these proofs make use of classical reasoning. The second proof is amenable to a
  constructive formalization as shown in the following function in \texttt{ARS-NewmansLemma.agda}:

@@

  We show in Subsection \ref{subsec:newnewman} a generalization of Newman's Lemma using a condition distilled from $\gRP$.
  The formalizations of the remaining theorems are all found in \texttt{ARS-Theorems.agda}.
\end{proof}

\begin{theorem}(Thm 1.2.2 in [Terese])
\begin{description}
  \item[($i$)] $\gCR \implies \gNPe \implies \gUN$
  \item[($ii$)] $\gWN \land \gUN \implies \gCR$
  \item[($iii$)] $\gCRs \implies \gCR$
\end{description}
\end{theorem}

\begin{proof}
    See, \\
    \verb|CR→NP : R isCR → R isNP₌|\\
    \verb|NP→UN : R isNP₌ → R isUN|\\
    \verb|ii : R isWN × R isUN → R isCR|\\
    \verb||\texttt{UN}$^{\to}$ \verb|∧WN→CR : R |\texttt{isUN}$^{\to}$ \verb| → R isWN → R isCR|\\
    \verb|iii : subcommutative R → R isCR|\\
    in \texttt{Theorems.agda}.
\end{proof}
\begin{proof}[Discussion.]
    In $\mathtt{UN^{→}∧WN→CR}$ a more general proof of ($ii$) is given using $\UNto$.
\end{proof}
\apcomment{Perhaps change the fourth implication to use $\times$ instead of $\to$ ??}

\begin{theorem}(Thm. 1.2.3 in [Terese])
  \begin{enumerate}[(i)]
    \item $\gWN \land \gUN \implies \gBP$
    \item $\gBP \land \gInc \implies \gSN$
    \item $\gWCR \land \gWN \land \gInc \implies \gSN$
    \item $\gCR \iff \gCP$ ($\gCR \implies \gCP$ only holds if the ARS is countable.)
  \end{enumerate}
  % \begin{description}
  %   \item[($i$)] $\gWN \land \gUN \implies \gBP$
  %   \item[($ii$)] $\gBP \land \gInc \implies \gSN$
  %   \item[($iii$)] $\gWCR \land \gWN \land \gInc \implies \gSN$
  %   \item[($iv$)] $\gCR \iff \gCP$ ($\gCR \implies \gCP$ only holds if the ARS is countable.)
  % \end{description}
\end{theorem}
\apcomment{Do the same for 1.2.2, and below!}

\begin{proof}
    See \texttt{Theorems.agda},\\
    \verb|i     : R isWN → R isUN → R isBP|\\
    $\mathtt{i^\to}$\hspace{5mm} \verb|: R isWN → R |\texttt{isUN}$^{\to}$ \verb|→ R isBP|\\
    \verb|iiSeq : R isWN → R isUN → R isRP → isWFseq- (~R R)|\\
    \verb|iii   : R isWN → R isWCR → R isRP- → dec (SN) → R isSN|\\
    \verb|iv    : R isCP → R isCR|
\end{proof}
\apcomment{Apply similar formatting elsewhere.}
\begin{proof}[Discussion.]\hfill
  \begin{description}
    \item[($i$)] In $\mathtt{i^\to}$ we again provide a slight improvement by using $\gUNto$ rather than $\gUN$.
    \item[($ii$)] This proof is inherently classical. In $\mathtt{iiSeq}$, we show that these properties imply $\isWFseqm$; establishing the implication to $\gSN$ would require additional classical assumptions. The definition and further discussion of $\isWFseqm$ are provided in Section~\ref{sec:Well-foundedness}.
    \item[($iii$)] In $\mathtt{iii}$, we replace $\gInc$ with the more general property $\gRPm$. We also assume that $\gSN$ is decidable, as this allows the construction of a witness for whether an element possesses the property $\SN$, which is essential for our proof.
    \item[($iv$)] We did not formalize $\gCR \to \gCP$ as it requires cardinality assumptions. \qedhere
  \end{description}
\end{proof}

% \begin{remark} \hfill
%     \begin{itemize}
%         \item In $\mathtt{i^\to}$ we again provide a slight improvement by using $\gUNto$ rather than $\gUN$.
%         \item ($ii$) is inherently classical. In $\mathtt{iiSeq}$, we show that these properties imply $\isWFseqm$; establishing the implication to $\gSN$ would require additional classical assumptions. The definition and further discussion of $\isWFseqm$ are provided in Section~\ref{sec:Well-foundedness}.
%         \item In $\mathtt{iii}$, we replace $\gInc$ with the more general property $\gRPm$. We also assume that $\gSN$ is decidable, as this allows the construction of a witness for whether an element possesses the property $\SN$, which is essential for our proof.
%         \item We did not formalize $\gCR \to \gCP$ as it requires cardinality assumptions.
%     \end{itemize}
% \end{remark}

### 2025-06-06 Relations/Paper/Summer_2025_TODO.tex
Paper revision

    \item (Review 1) Improve fonts for constant and text (refering to a definition on line 258). General formatting imrpovement for how we set out defns.

@@

    \item (Review 2) Not convinced by our definitions, as they are imprecise and lack quantifiers. 
    \item (Review 2) Issue with definition of R|=RP on line 188. Use of bounded is ambiguous. We should clarify which use of bounded.
    \item (Our notes) Improve proposition 11 on line 194, we want to make the clame that RP- $\to$ RP, and either have a one line proof or refer to the proof in Agda. 
    \item (Review 2) Section 3 is "full of repetitions and paraphrases between the statements in natural language and corresponding formalization". We can try and think of a way to get it to flow better, we do want to provide natural language summaries of what is in the code!
    \item [\done] (Review 2) Struggled to understand the implications tables. We should add a reminder that bold font is global. We should rethink the tables generally to make them clearer. Don't comma seperate and don't denote counterexamples as CE-XX.
    \item (Review 2) Struggled to find the relevant agda code for the various definitions. We should find a way to better link the two.
    \item (Review 2) Clarify Notation 7 on line 159: "Choose a uniform notation, X or NP, but not both".
    \item [\done] (we think) (Review 2) "l 169: An annoying notational collision between the two arrows -> at different heights."I'm not too sure what they're referring to, though there could be some ambiguity with the arrows here, especially with use of $UN\to$.

@@

\section*{Week starting 2nd June}
\begin{todolist}
  \item Should we add quantification to each definition(I think quantification for $a$ is already implied)? E.g.\\ 
  \item [\ule{$a \in \WCR_R$}] \emph{$a$ is weakly Church-Rosser (or weakly confluent)} if $c \bstep a \rstep b \implies c \mstep d \bmstep b$.\\
   Transforms to 
  \item[\ule{$a \in \WCR_R$}] \emph{$a$ is weakly Church-Rosser (or weakly confluent)} if $c \bstep a \rstep b \implies c \mstep d \bmstep b$ for all $b, c, d \in A$. 
  \item For showing inc -> RP we're currently just saying "This is immediate". Is this sufficient? We did have a proof previously but I think we removed it for space reasons. 
\end{todolist}
\end{document}

>>>>

    \item [\done] Improve fonts for constant and text (refering to a definition on line 258). General formatting imrpovement for how we set out defns.

@@

    \item [\done] Not convinced by our definitions, as they are imprecise and lack quantifiers.
    \item [\done] Issue with definition of R|=RP on line 188. Use of bounded is ambiguous. We should clarify which use of bounded.
    \item [\done] Improve proposition 11 on line 194, we want to make the clame that RP- $\to$ RP, and either have a one line proof or refer to the proof in Agda.
    \item [\done] Section 3 is "full of repetitions and paraphrases between the statements in natural language and corresponding formalization". We can try and think of a way to get it to flow better, we do want to provide natural language summaries of what is in the code!
    \item [\done] (Review 2) Struggled to understand the implications tables. We should add a reminder that bold font is global. We should rethink the tables generally to make them clearer. \apcomment{Don't comma seperate and don't denote counterexamples as CE-XX. ?}
    \item [\done] Struggled to find the relevant agda code for the various definitions. We should find a way to better link the two.
    \item [\done] Clarify Notation 7 on line 159: "Choose a uniform notation, X or NP, but not both".
    \item [\done] l 169: An annoying notational collision between the two arrows -> at different heights."I'm not too sure what they're referring to, though there could be some ambiguity with the arrows here, especially with use of $UN\to$.

@@

\section*{Week starting 2nd June}
\begin{todolist}
  \item Should we add quantification to each definition(I think quantification for $a$ is already implied)? E.g.\\
  \item [\ule{$a \in \WCR_R$}] \emph{$a$ is weakly Church-Rosser (or weakly confluent)} if $c \bstep a \rstep b \implies c \mstep d \bmstep b$.\\
   Transforms to
  \item[\ule{$a \in \WCR_R$}] \emph{$a$ is weakly Church-Rosser (or weakly confluent)} if $c \bstep a \rstep b \implies c \mstep d \bmstep b$ for all $b, c, d \in A$.
  \item For showing inc -> RP we're currently just saying "This is immediate". Is this sufficient? We did have a proof previously but I think we removed it for space reasons.
  \item Remove "desired implications" and similar organizational issues in the code (general review)
  \item Consider defining FB in terms of existing list library; especially in the rebooted repo,
  we shouldn't use manual List type just for this.
  \item Move all definitions and properties from Section 2.3 to Seq.agda, except Inc.
  \item Revise section 3.1
\end{todolist}
\end{document}
